---
title: "txseq: Differential gene expresssion analysis report"
header-includes: #allows you to add in your own Latex packages
- \usepackage{float} #use the 'float' package
- \floatplacement{figure}{H} #make every figure with caption = h
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: spacelab
    highlight: tango
    fig_caption: true
    code_folding: hide
urlcolor: blue
---

```{r deseq2-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.extra = '')
```


```{r message=FALSE}
library(ggplot2)
suppressPackageStartupMessages(library(circlize))
library(gridExtra)
library(DESeq2)
library(reshape2)
library(dplyr)
library(vsn)
library(gplots)
library(knitr)
library(openxlsx)
library(genefilter)
library(pander)
library(ggrepel)
library(ComplexHeatmap)
library(txseq)
```


```{r include=FALSE, echo=FALSE}
# Report configuration

setwd("~/work/txseq/hsc_example/reports/")

# Set the location of the txseq project run
txseq_run = "~/work/txseq/hsc_example"

# If you have used non-standard folder names you may need
# to also edit these paths.
salmon_db =  file.path(txseq_run, "salmon/csvdb")

# Set the location_function and fit type
location_function = "median"
fit_type = "local"

# Threshold for significance
p_value_threshold <- 0.05

# Effect size threshold
abs_fc_threshold <- 1.5
```


```{r include=FALSE, echo=FALSE}

# Load the TPM data and gene annotations

# load annotation data
statement = 'select distinct gene_id, gene_name, gene_biotype from transcript_info'
ann_df = fetch_DataFrame(statement, salmon_db)
ann_df$gene_name[is.na(ann_df$gene_name)] <- ann_df$gene_id[is.na(ann_df$gene_name)]
rownames(ann_df) <- ann_df$gene_id


# load in the tpms (for heatmaps)
tpms = fetch_DataFrame("select * from salmon_genes_tpms", salmon_db)
rownames(tpms) <- tpms$gene_id
tpms$gene_id <- NULL
tpms <- upperQuartileNormalise(tpms)
```



```{r include=FALSE, echo=FALSE}

# Load the expression data, metadata and gene annotations

# ** The workflow shown here assumes that we have counts from fragments of full length RNA
#    and not e.g. 3' lexogen data. **
#
# From the DESeq2 vignette:
#   If you have 3’ tagged RNA-seq data, then correcting the counts for gene length will induce a bias in your analysis, 
#   because the counts do not have length bias. Instead of using the 
#   default full-transcript-length pipeline, we recommend to use the original counts, e.g. txi$counts as a counts matrix, 
#   e.g. providing to DESeqDataSetFromMatrix or to the edgeR or # limma functions without calculating an offset and 
#   without using countsFromAbundance.

# Load in the tximeta file containing the count data from Salmon
edata <- readRDS(file.path(txseq_run,"salmon/tximeta.dir/tximeta.RDS"))

# Extract the gene-level data
gse <- edata$gse


# set a grouping variable which combines cell type and condition
colData(gse)$group <- factor(paste(colData(gse)$cell_type,colData(gse)$condition,sep="_"))

# Make the DESeq2 dataset
dds <- DESeqDataSet(gse,
                    design=formula("~group"))


# Use a txseq R library function to pre-filter the data
dds <- filterDESeqDatasetByGroup(dds,
                                 grouping_var="group",
                                 min_replicates=2,
                                 min_counts=5)

```


# Introduction

Differential expression testing was performed using the [DESeq2](http://www.bioconductor.org/packages/release/bioc/html/DESeq2.html) package.

# DESeq2 parameter selection

## Selection of method for estimating the location of the size factors

The figure shows a comparison of median and shorth methods for size-factor location estimation.

```{r de-location,  fig.height=6, fig.width=10,  echo=FALSE, message=FALSE}

# compare the median and shorth methods for estimating location
d <- counts(dds)

# exclude genes with 0 counts
d <- d[apply(d,1,function(x) min(x) > 0),]

# geometric mean function - note that missing values/nas are not expected!
gm <- function(x) { exp(sum(log(x)) / length(x)) }

# get the geometric means
gms <- apply(d,1, gm)

# get the ratios
r <- d/gms



# estimate the locations
meds <- apply(r,2,median)
shs <- apply(r,2,shorth)
loc_df <- data.frame(median=meds, shorth=shs,sample=names(meds))

# plot the fits
mr <- melt(r)
colnames(mr) <- c("gene_id","sample","value")

gp <- ggplot(mr, aes(value)) + geom_freqpoly()
gp <- gp + geom_vline(data=loc_df, aes(xintercept=median),color="red")
gp <- gp + geom_vline(data=loc_df, aes(xintercept=shorth),color="blue")
gp <- gp + scale_x_continuous(trans="log2")#, limits=c(0.25,4))
gp <- gp + facet_wrap(~sample)
gp <- gp + xlab("Per-gene size factors. Red line: median, Blue line: shorth")
gp <- gp + theme_light()

print(gp)

```

Based on these plots, we choose to use the median as the best estimator of the location.


## Selection of method for fitting expression-level variance relationship

The figure shows a comparison of local and parametric fits for modelling the expression level-variance relationship

```{r de-fittype, fig.height=4, fig.width=10,  echo=FALSE, message=FALSE}

dds_test <- dds

par(mfrow=c(1,2))

    for(fitType in c("local","parametric"))
    {
        dds_test <- estimateSizeFactors(dds_test, locfunc=match.fun(location_function))
        dds_test <- estimateDispersions(dds_test, fitType=fitType)

        plot_title = paste0("locfunc: ",location_function,", fit-type: ",fitType)
        plotDispEsts(dds_test, main=plot_title)

    }
```

Based on the plots, we choose to use a local fit.


# Identification of genes differentially expressed in multi-potent progenitors

Next we run a test for genes differentially expressed between multipotent progenitors in steady state vs curdlan treatment. 

```{r de-run-analysis, echo=FALSE, message=FALSE}

# Set the correct reference level
dds$group <- relevel(dds$group, ref = "lthsc_sscontrol")

# Estimate the size factors, dispersions and run a Wald test
dds <- estimateSizeFactors(dds, locfunc=match.fun(location_function))
dds <- estimateDispersions(dds, fitType=fit_type)
dds <- nbinomWaldTest(dds)

# Check the results names
resultsNames(dds)

# Shrink the fold-changes for visualisation and gene-ranking purposes
resLFC <- lfcShrink(dds, coef="group_lthsc_curdlan_vs_lthsc_sscontrol", type="apeglm")

res <- resLFC

res$gene_name = ann_df[rownames(res), "gene_name"]
res$gene_id <- rownames(res)


```

## Data normalisation

The normalisation of the data and model fit is first checked.

```{r fig.width=6, fig.height=5,  echo=F}

# If using DESeqDataSetFromMatrix, e.g. without length correction,
# simply extract the size factors with:
#    SFs <- sizeFactors(dds)
SFs <- colMeans(normalizationFactors(dds))
size_df <- as.data.frame(SFs, drop = F)
s <- rownames(size_df)
size_df$samples <- factor(s, levels = s[order(size_df$SFs)])
gp <- ggplot(size_df, aes(samples, SFs)) + geom_bar(stat = "identity")
gp <- gp + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("")
gp <- gp + ylab("normalisation/size factor") 
sf_plot <- gp

# inspect the normalisation
norm_plots <- plotDensities(counts(dds,normalized=T), 
                            colData(dds),
                            "condition", "normalised counts (log2)")

gps <- list(norm_plots$den, norm_plots$bp, sf_plot)
grid.arrange(grobs=gps, nrows=2,ncol=2)

```


## Model fit and independent filtering plots

```{r fig.width=10, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}

par(mfrow=c(1,2))

# plot the dispersion-expression level model fit
plotDispEsts(dds)

# plot the independent filtering model fit
plot(metadata(res)$filterNumRej, type = "b", ylab = "number of rejections", 
        xlab = "quantiles of filter")
lines(metadata(res)$lo.fit, col = "red")
abline(v = metadata(res)$filterTheta)
```

## Numbers of DE genes identified

```{r include=FALSE, echo=FALSE}



tt <- res
tt <- tt[!is.na(tt$padj),]
write.table(tt,"tt.salmon.tsv",col.names=T,row.names=F,sep="\t",quote=F)

nup <- dim(tt[tt$padj< p_value_threshold & tt$log2FoldChange> log2(abs_fc_threshold),])[1]
ndown <- dim(tt[tt$padj< p_value_threshold & tt$log2FoldChange< -log2(abs_fc_threshold),])[1]

summary_mat <- matrix(c("deseq_results", nup, ndown),nrow=1)
summary <- data.frame(summary_mat)

colnames(summary) <- c("comparison","n_up","n_down")
rownames(summary) <- summary$comparison
summary$comparison <- NULL

kable(summary,caption=paste(
	"Numbers of differentially expressed genes (padj < ",
	 p_value_threshold,
	", fold change > ",
	abs_fc_threshold,
	")"
))
```


## MA and volcano plots of DE genes

The figure show the differentially expressed genes visualised on an MA plot (left) and a volcano plot (right).

```{r fig.width=10, fig.height=5, echo=FALSE}

# get the identifiers of genes to highlight.
interesting_genes <- get_interesting(tt,
                                    p_value_threshold=p_value_threshold,
                                    n=10,
                                    p_col="padj",
                                    fc_col="log2FoldChange")

par(mfrow=c(1,2))
deplots <- de_plots(data.frame(tt),
               interesting_genes=interesting_genes,
               abs_fc_threshold=abs_fc_threshold)

grid.arrange(grobs=(list(deplots$ma, deplots$volcano)), ncol=2)
 
```

## Heatmap of top differentially expressed genes

```{r , fig.width=4, fig.height=10, echo=FALSE, warning=FALSE, message=FALSE}

# Extract a list of the samples included in the contrast of interest
samples <- colData(dds)$sample_id[colData(dds)$group %in% c("lthsc_sscontrol","lthsc_curdlan")]
samples <- samples[order(samples)]

data <- log2(tpms[interesting_genes, samples]+1)
rownames(data) <- ann_df[rownames(data),"gene_name"]

col_fun = colorRamp2(range(data), hcl_palette = "YlOrRd", reverse = TRUE)
Heatmap(data,
        col=col_fun,
        row_names_gp = gpar(fontsize = 10),
        heatmap_legend_param = list(title = "log2(TPM+1)"),)


```



## Recreating a figure from the publication

As a sanity check, we recreate Figure panel 2C from Regan-Komito et al. The plot looks very similar, despite the use of a different quantification approach.

```{r , fig.width=3, fig.height=4, echo=FALSE, warning=FALSE, message=FALSE}

selected_genes <- c("Selp","Csf2rb","Cd14","Lbp","Vldlr","Cdk6","Cd53","Plac8")

selected_gene_ids <- rownames(ann_df)[ann_df$gene_name %in% selected_genes]

# Extract a list of the samples included in the contrast of interest
data <- log2(tpms[selected_gene_ids , samples]+1)
rownames(data) <- ann_df[rownames(data),"gene_name"]

data <- data[selected_genes,c("sscontrol_bm_lthsc_R1","sscontrol_bm_lthsc_R2","sscontrol_bm_lthsc_R3",
                              "curdlan_bm_lthsc_R1","curdlan_bm_lthsc_R2","curdlan_bm_lthsc_R3")]

#col_fun = colorRamp2(range(data), hcl_palette = "YlOrRd", reverse = TRUE)
Heatmap(t(scale(t(data))),
        cluster_rows = F,
        cluster_columns = F,
        #col=col_fun,
        row_names_gp = gpar(fontsize = 10),
        heatmap_legend_param = list(title = "log2(TPM+1)"),)


```


# Saving the results as an Excel file

Finally we use the openxlsx library to save the results as an excel file. This is useful, e.g. for preparation of supplementary tables.

```{r de-write-tables, echo=FALSE, message=FALSE, warning=FALSE}

# save the result in an xlsx tables.

xlsx_file = "deseq2_results.xlsx"
wb <- write.xlsx(list(results=data.frame(tt)), file = xlsx_file, asTable=TRUE)
saveWorkbook(wb, xlsx_file, overwrite = TRUE)

```

# Session Information

```{r}
sessionInfo()
```



